---
title: "CENG 3516 STATISTICAL COMPUTING"
author: "Dr. Zeynep Filiz EREN DOGU"
date: "Feb 28,2021"
output:
  html_document: default
  pdf_document: default
subtitle: Lecture 1 - Introduction and Basics
font-family: Gill Sans
---

###In-class exercise: Hello world!
```{r}
"Hello world!"
print("Hello world!")
```

###Setting and getting working directory
```{r}
#getwd()
#setwd()
```

###Basics: the class in a nutshell
- Everything we'll do comes down to applying **functions** to **data**

- **Data**:  things like 7, "seven", $7.000$, the matrix $\left[ \begin{array}{ccc} 7 & 7 & 7 \\ 7 & 7 & 7\end{array}\right]$

- **Functions**: things like $\log{}$, $+$, $<$, $\mod{}$, `mean`, `sum`, `data`

> A function is a process which turns input objects (**arguments**) into an output object (**return value**)

###Data building blocks

You'll encounter different kinds of data types

- **Booleans** Direct binary values: `TRUE` or `FALSE` in R
- **Integers**: whole numbers (positive, negative or zero)
- **Characters** fixed-length blocks of bits, with special coding
**strings** = sequences of characters
- **Floating point numbers**: a fraction (with a finite number of bits) times an exponent, like $1.87 \times {10}^{6}$
- **Missing or ill-defined values**: `NA`, `NaN`, etc.

###Operators (functions)
You can use R as a calculator

Command|Description
----------|-------------
`+,-,*,\` | add, subtract, multiply, divide
`^`       | raise to the power of
`%%`      | remainder after division (ex: `8 %% 3 = 2`)
`( )`     | change the order of operations
`log(), exp()` | logarithms and exponents (ex: `log(10) = 2.302`)
`sqrt()` | square root
`round()` | round to the nearest whole number (ex: `round(2.3) = 2`)
`floor(), ceiling()` | round down or round up 
`abs()` | absolute value

```{r}
7 + 5 # Addition
7 - 5 # Subtraction
7 * 5 # Multiplication
7 ^ 5 # Exponentiation
"Hello world."
exp(1)
cos(3.141593)
log2(1)
```
###

```{r}
7 / 5 # Division
7 %% 5 # Modulus
7 %/% 5 # Integer division 
```
###Operators cont'd.
**Comparisons** are also binary operators; they take two objects, like numbers, and give a Boolean

```{r}
7 > 5
7 < 5
7 >= 7
7 <= 5
```

###Operators cont'd.
```{r}
7 == 5
7 != 5
```

###Boolean operators
Basically "and" and "or":
```{r}
(5 > 7) & (6*7 == 42)
(5 > 7) | (6*7 == 42)
```

###  More on types
- `typeof()` function returns the type

- `is.`_foo_`()` functions return Booleans for whether the argument is of type _foo_

- `as.`_foo_`()` (tries to) "cast" its argument to type _foo_ --- to translate it sensibly into a _foo_-type value

<small>**Special case**: `as.factor()` will be important later for telling R when numbers are actually encodings and not numeric values.  (E.g., 1 = High school grad; 2 = College grad; 3 = Postgrad) </small>
###
```{r}
typeof(7)
is.numeric(7)
is.na(7)
```
###  
```{r}
is.character(7)
is.character("7")
is.character("seven")
is.na("seven")
```
  
###Variables
We can give names to data objects; these give us **variables**
A few variables are built in:
```{r}
pi
letters
## months in English
month.abb
month.name
## months in your current locale
format(ISOdate(2000, 1:12, 1), "%B")
format(ISOdate(2000, 1:12, 1), "%b")
```

Variables can be arguments to functions or operators, just like constants:
```{r}
pi*10
cos(pi)
pi - 4*(4*atan(1/5) - atan(1/239))
```

###Assignment operator
Most variables are created with the **assignment operator**, `<-` or `=`  
```{r}
x <- 2
x = 2
x
x ^ x
x ^ 2
```

###Assignment operator
Most variables are created with the **assignment operator**, `<-` or `=`  
```{r}
hours <- 12
hours
days = 2.5
hours * days
```

###
The assignment operator also changes values:
```{r}
total.hours<- hours * days
total.hours
hourly.wage <- 45
payment<-total.hours*hourly.wage
```

###
- Using names and variables makes code: easier to design, easier to debug, less prone to bugs, easier to improve,  and easier for others to read

- Avoid "magic constants"; use named variables

- Use descriptive variable names and dots between names, variable names are case sensitive
  - Good: `num.students <- 29`
  - Bad: `ns <- 29 `
  
###The workspace
What names have you defined values for?
```{r}
ls()
```

Removing variables:
```{r}
rm("hourly.wage") ###remove variables
ls()
```

###First data structure: vectors
- Group related data values into one object, a **data structure**
- A **vector** is a sequence of values, all of the same type
- `c()` function returns a vector containing all its arguments in order
```{r}
x <- c(1:10)
x
x ^ x
dim(x) <- c(2,5)
x <- c("Hello","world","!")
x
x <- c(TRUE, TRUE, FALSE, FALSE, TRUE, FALSE, TRUE)
x
x <- list("R","12345",FALSE)
x
students <- c("Meltem", "Umut", "Olcay", "Busra", "Hayriye")
midterm <- c(80, 90, 93, 82, 95)
ls()
```
###Using rep() and seq() functions
```{r}
Example1 <- seq(2:50)
Example1 
Example1 <- seq(1, 9, by = 2)
Example1 
Example1 <-seq(1.575, 5.125, by = 0.05)
Example1 
Example1 <- rep(1:10,2)
Example1 
Example1 <- rep(1:2,c(10,15))
Example1 
Example1 <- rep(1:2,each=10)
Example1 
Example1 <- rep(1:2,c(10,10))
Example1 
Example2 <- log(Example1)
Example2
ls()
```

###Indexing
- `vec[1]` is the first element, `vec[4]` is the 4th element of `vec`
```{r}
students
students[4]
```
- `vec[-4]` is a vector containing all but the fourth element
```{r}
students[-4]
```

###Second data structure: matrices
- A **matrix** is a set of vectors
- `matrix()` function returns a matrix containing all its arguments in order
```{r}
matrix1<-matrix(1:10, 2, 5)
matrix1
matrix1[2,2]
matrix1[,2]
matrix1[2,]
matrix2<-matrix(0, 5, 2)
matrix2
matrix3<-matrix(NA, 5, 2)
matrix3
class(matrix3)
attributes(matrix1)
colnames(matrix1) <- c("A","B","C","D","E")
matrix4<-matrix(data=c(1,2,3,4,5,6,7,8,9,10,11,12),nrow=3,ncol=4)
matrix4
matrix5<-matrix(1:12,nrow=3,byrow=T)
matrix5
######Other ways of creating matrices
matrix6<-cbind(A=1:4,B=5:8,C=9:12)
matrix6
matrix7<-rbind(A=1:4,B=5:8,C=9:12)
matrix7
```

###Vector arithmetic
Operators apply to vectors "pairwise" or "elementwise":
```{r}
final <- c(88, 84, 95, 82, 91) # Final exam scores
midterm # Midterm exam scores
midterm + final # Sum of midterm and final scores
((0.5)*midterm + (0.5)*final) # Average exam score
course.grades <- (0.3)*midterm + (0.7)*final # Final course grade
course.grades
```

###Pairwise comparisons
Is the final score higher than the midterm score?
```{r}
midterm 
final
final > midterm
```

Boolean operators can be applied elementwise:
```{r}
(final < midterm) & (midterm > 80)
```

###Functions on vectors
Command | Description
--------|------------
`sum(vec)` | sums up all the elements of `vec`
`mean(vec)` | mean of `vec`
`median(vec)` | median of `vec`
`min(vec), max(vec)` | the largest or smallest element of `vec`
`sd(vec), var(vec)` | the standard deviation and variance of `vec`
`length(vec)` | the number of elements in `vec`
`pmax(vec1, vec2), pmin(vec1, vec2)` | example: `pmax(quiz1, quiz2)` returns the higher of quiz 1 and quiz 2 for each student
`sort(vec)` | returns the `vec` in sorted order
`order(vec)` | returns the index that sorts the vector `vec`
`unique(vec)` | lists the unique elements of `vec`
`summary(vec)` | gives a five-number summary  
`any(vec), all(vec)` | useful on Boolean vectors

###Functions on vectors
```{r}
course.grades
sum(midterm,final)/2
mean(course.grades) # mean grade
median(course.grades)
sd(course.grades) # grade standard deviation
```

###More functions on vectors
```{r}
sort(course.grades)
max(course.grades) # highest course grade
min(course.grades) # lowest course grade
```
###Referencing elements of vectors
```{r}  
students
```
  
  Vector of indices:
```{r}
students[c(2,4)]
```

Vector of negative indices
```{r}
students[c(-1,-3)]
```

###More referencing

`which()` returns the `TRUE` indexes of a Boolean vector:
```{r}
course.grades
a.threshold <- 90 # A grade = 90% or higher
course.grades >= a.threshold # vector of booleans
a.students <- which(course.grades >= a.threshold) # Applying which() 
a.students
students[a.students] # Names of A students
```

###Named components
  
You can give names to elements or components of vectors
```{r}
students
names(course.grades) <- students # Assign names to the grades
names(course.grades)
course.grades[c("Meltem", "Umut", "Olcay")] # Get final grades for 3 students
```
Note the labels in what R prints; these are not actually part of the value

###Useful RStudio tips
Keystroke | Description
----------|-------------
`<tab>` | autocompletes commands and filenames, and lists arguments for functions. Highly useful!
`<up>` | cycle through previous commands in the console prompt
`<ctrl-up>` | lists history of previous commands matching an unfinished one
`<ctrl-enter>` | paste current line from source window to console. Good for trying things out ideas from a source file.
`<ESC>` | as mentioned, abort an unfinished command and get out of the + prompt

<br>
