


###Agenda

- Summaries with the aggregate() function
- Standard graphics

###Getting started: birthwt data set-Risk Factors Associated with Low Infant Birth Weight
- We're going to start by operating on the `birthwt` dataset from the MASS library

- Description: The birthwt data frame has 189 rows and 10 columns. The data were collected at Baystate Medical Center, Springfield, Mass during 1986.

- Usage: birthwt

- Format: This data frame contains the following columns:

  low: indicator of birth weight less than 2.5 kg.

  age: mother's age in years.

  lwt: mother's weight in pounds at last menstrual period.

  race: mother's race (1 = white, 2 = black, 3 = other).

  smoke: smoking status during pregnancy.

  ptl: number of previous premature labours.

  ht: history of hypertension.

  ui: presence of uterine irritability.

  ftv: number of physician visits during the first trimester.

  bwt: birth weight in grams.

```{r}
library(MASS)
str(birthwt)
```
###Renaming the variables
- The dataset doesn't come with very descriptive variable names

- Better names can be used for better understanding

```{r}
colnames(birthwt) 

# The default names are not very descriptive

colnames(birthwt) <- c("birthwt.below.2500", "mother.age", "mother.weight", 
    "race", "mother.smokes", "previous.prem.labor", "hypertension", "uterine.irr", 
    "physician.visits", "birthwt.grams")

# Better names!
```

###Renaming the factors

- All the factors are currently represented as integers

- Let's use the `transform()` and `mapvalues()` functions to convert variables to factors and give the factors more meaningful levels 
```{r}
library(plyr)
birthwt <- transform(birthwt, 
            race = as.factor(mapvalues(race, c(1, 2, 3), 
                              c("white","black", "other"))),
            mother.smokes = as.factor(mapvalues(mother.smokes, 
                              c(0,1), c("no", "yes"))),
            hypertension = as.factor(mapvalues(hypertension, 
                              c(0,1), c("no", "yes"))),
            uterine.irr = as.factor(mapvalues(uterine.irr, 
                              c(0,1), c("no", "yes"))),
            birthwt.below.2500 = as.factor(mapvalues(birthwt.below.2500,
                              c(0,1), c("no", "yes")))
            )
```

###Summary of the data
- Now that things are coded correctly, we can look at an overall summary

```{r}
summary(birthwt)
mean(birthwt$birthwt.grams)
# Calculate mean for smokers and nonsmokers
sd(birthwt$birthwt.grams)
# Calculate standard deviation for smokers and nonsmokers
fivenum(birthwt$birthwt.grams)
# Calculate five number summary for smokers and nonsmokers
```

###A simple table
- Let's use the `tapply()` function to see what the average birthweight looks like when broken down by race and smoking status

```{r}
with(birthwt, tapply(birthwt.grams, INDEX = list(race, mother.smokes), FUN = mean)) 
with(birthwt, tapply(birthwt.grams, INDEX = list(race, mother.smokes), FUN = sd)) 
with(birthwt, tapply(birthwt.grams, INDEX = list(race, mother.smokes), FUN = median)) 
with(birthwt, tapply(birthwt.grams, INDEX = list(race, mother.smokes), FUN = IQR)) 
```

###aggregate() function
- Let's first recall what `tapply()` does

- Command: `tapply(X, INDEX, FUN)` 
    - Applies `FUN` to `X` grouped by factors in `INDEX`
  
- **`aggregate()`** performs a similar operation, but presents the results in a form that is at times more convenient

- There are many ways to call the `aggregate()` function

- Analog of `tapply` call: `aggregate(X, by, FUN)` 
    - Here, `by` is exactly like `INDEX`
    
###Example: tapply vs aggregate
```{r}
library(MASS)
with(birthwt, tapply(birthwt.grams, INDEX = list(race, mother.smokes), FUN = mean)) # tapply
with(birthwt, aggregate(birthwt.grams, by = list(race, mother.smokes), FUN = mean)) # aggregate
# What the average birthweight looks like when broken down by race and hypertention?
```

###Example: different syntax
- Here's a convenient alternative way to call `aggregate`

- It uses the R `formula` syntax, which we'll learn more about when we discuss regression

```{r}
aggregate(birthwt.grams ~ race + mother.smokes, FUN=mean, data=birthwt)
```

###A closer look at low birth weight 

```{r}
table(birthwt$race,birthwt$mother.smokes)
weight.smoke.tbl <- with(birthwt, table(birthwt.below.2500, mother.smokes))
weight.smoke.tbl
# Create the first crosstab using with() function
```

- Is the mother's age correlated with birth weight?

```{r}
par(mfrow = c(2,2))
plot(birthwt$birthwt.grams)
plot(birthwt$birthwt.grams,birthwt$mother.age)
plot(birthwt$birthwt.grams,birthwt$mother.weight)
cor(birthwt$birthwt.grams,birthwt$mother.weight)
cor(birthwt$birthwt.grams,birthwt$mother.age)  # Calculate correlation
# Calculate the correlation coefficient between birthweight and mother's age using with() function
```

- Does this change when we account for smoking status?

```{r}
par(mfrow = c(1,1))
plot(birthwt$birthwt.grams[birthwt$mother.smokes == "yes"], birthwt$mother.age[birthwt$mother.smokes == "yes"])
# Create previous plot using with() function
# Calculate the correlation coefficient for the previous plot using with() function
# Compute the summary statistics and correlation for nonsmokers and create a plot for it. 
```

###Faster way: by() function
- Think of the `by(data, INDICES, FUN)` function as a `tapply()` function that operates on data frames instead of just vectors

- When using `tapply(X, INDEX, FUN)`, `X` is generally a numeric vector

- To calculate correlations, we need to allow `X` to be a data frame or matrix

```{r}
by(data = birthwt, 
   INDICES = birthwt[,"mother.smokes"], 
   FUN = summary)
by(data = birthwt$birthwt.below.2500, 
   INDICES = birthwt["mother.smokes"], 
   FUN = summary)
by(data = birthwt$mother.weight, 
   INDICES = birthwt["mother.smokes"], 
   FUN = mean)
```



###Standard graphics in R

#### Single-variable plots

Let's continue with the `birthwt` data from the `MASS` library.

Here are some basic single-variable plots.

```{r, fig.height = 7, fig.align='center'}
par(mfrow = c(2,4)) # Display plots in a single 2 x 2 figure 
plot(birthwt$mother.age)
with(birthwt, hist(mother.age))
with(birthwt, boxplot(mother.age))
plot(birthwt$mother.smokes)
plot(birthwt$birthwt.grams)
plot(birthwt$birthwt.grams,birthwt$mother.weight)
qqnorm(birthwt$birthwt.grams)
qqline(birthwt$birthwt.grams)
qqnorm(birthwt$mother.weight)
qqline(birthwt$mother.weight)
```

Note that the result of calling `plot(x, ...)` varies depending on what `x` is.  
    - When `x` is *numeric*, you get a plot showing the value of `x` at every index.  
    - When `x` is a *factor*, you get a bar plot of counts for every level


Let's add more information to the smoking bar plot, and also change the color by setting the `col` option.

```{r, fig.height=5, fig.width=5, fig.align='center'}
par(mfrow = c(1,1))
plot(birthwt$mother.smokes, 
     main = "Mothers Who Smoked In Pregnancy", 
     xlab = "Smoking during pregnancy", 
     ylab = "Count of Mothers",
     col = "lightgrey")
boxplot(birthwt$birthwt.grams~birthwt$hypertension)
# Create a boxplot for birthweights conditioned on smoking
# Create a boxplot for birthweights conditioned on physician visits
```

### (much) better graphics with ggplot2

#### Introduction to ggplot2

ggplot2 has a slightly steeper learning curve than the base graphics functions, but it also generally produces far better and more easily customizable graphics.  

There are two basic calls in ggplot: 

- `qplot(x, y, ..., data)`: a "quick-plot" routine, which essentially replaces the base `plot()`
- `ggplot(data, aes(x, y, ...), ...)`: defines a graphics object from which plots can be generated, along with *aesthetic mappings* that specify how variables are mapped to visual properties.  

```{r}
library(ggplot2)
```

#### plot vs qplot

Here's how the default scatterplots look in ggplot compared to the base graphics.  We'll illustrate things by continuing to use the birthwt data from the `MASS` library.

```{r, fig.align='center', fig.height=3, fig.width=4 , fig.align='center'}
with(birthwt, plot(mother.age, birthwt.grams))  # Base graphics 
qplot(x=mother.age, y=birthwt.grams, data=birthwt)  # using qplot from ggplot2
```

Remember how it took us some effort last time to add color coding, use different plotting characters, and add a legend?  Here's the `qplot` call that does it all in one simple line.

```{r, fig.align='center', fig.height=4, fig.width=5, fig.align='center'}
qplot(x=mother.age, y=birthwt.grams, data=birthwt,
      color = mother.smokes,
      shape = mother.smokes,
      xlab = "Mother's age (years)",
      ylab = "Baby's birthweight (grams)"
      ) 
```

This way you won't run into problems of accidentally producing the wrong legend.  The legend is produced based on the `colour` and `shape` argument that you pass in.  (Note: `color` and `colour` have the same effect.  )

#### ggplot function

```{r fig.width=10, fig.height=4, dpi=70, cache=TRUE, fig.align='center'}
ggplot(data=birthwt, aes(x=birthwt.grams, y=mother.age))
birthwt.plot <- ggplot(data=birthwt, aes(x=birthwt.grams, y=mother.age))
birthwt.plot + geom_point()
```

Let's take a step back and try to understand the ggplot syntax.  

1) The first thing we did was to define a graphics object, `birthwt.plot`.  This definition told R that we're using the `birthwt` data, and that we want to display `carat` on the x-axis, and `price` on the y-axis.  

2) We then called `birthwt.plot + geom_point()` to get a scatterplot.

The arguments passed to `aes()` are called **mappings**.  Mappings specify what variables are used for what purpose.  When you use `geom_point()` in the second line, it pulls `x`, `y`, `colour`, `size`, etc., from the **mappings** specified in the `ggplot()` command.  

You can also specify some arguments to `geom_point` directly if you want to specify them for each plot separately instead of pre-specifying a default.

Here we shrink the points to a smaller size, and use the `alpha` argument to make the points transparent.

```{r fig.width=10, fig.height=4, dpi=70, cache=TRUE, fig.align='center'}
birthwt.plot + geom_point(size = 0.7, alpha = 0.3)
```

If we wanted to let point color depend on the color indicator of the diamond, we could do so in the following way.

```{r fig.width=10, fig.height=6, dpi=70, cache=TRUE, fig.align='center'}
birthwt.plot <- ggplot(data=birthwt, aes(x=birthwt.grams, y=mother.age, colour = mother.smokes))
birthwt.plot + geom_point()
```

We can change colors by specifying a different color palette.  Here's how we can switch to the `cbPalette` we saw last class.

```{r fig.width=10, fig.height=6, dpi=70, cache=TRUE, fig.align='center'}
cPalette <- c("#999999", "navyblue")
birthwt.plot <- ggplot(data=birthwt, aes(x=birthwt.grams, y=mother.age, colour = mother.smokes))
birthwt.plot + geom_point() + scale_colour_manual(values=cPalette)
```

To make the scatterplot look more typical, we can sometimes switch to logarithmic coordinate axis spacing.

```{r, eval = FALSE}
birthwt.plot + geom_point() +
  coord_trans(x = "log10", y = "log10")
```

#### Conditional plots

We can create plots showing the relationship between variables across different values of a factor.  For instance, here's a scatterplot showing how birth weight varies with hypertension, conditioned on color.  It's created using the `facet_wrap(~ factor1 + factor2 + ... + factorn)` command.

```{r, fig.width=12, fig.height=6, dpi=70, cache=TRUE}
birthwt.plot <- ggplot(data=birthwt, aes(x=birthwt.grams, y=mother.age))
birthwt.plot + geom_point() + facet_wrap(~ mother.smokes)

birthwt.plot <- ggplot(data=birthwt, aes(x=birthwt.grams, y=mother.age, colour = mother.smokes))
birthwt.plot + geom_point() + facet_wrap(~ hypertension)
```

`ggplot` can create a lot of different kinds of plots, just like lattice.  Here are some examples.

Function | Description
---------------------|------------------------------------------------
`geom_point(...)` | Points, i.e., scatterplot
`geom_bar(...)` | Bar chart
`geom_line(...)` | Line chart
`geom_boxplot(...)` | Boxplot
`geom_violin(...)` | Violin plot
`geom_density(...)` | Density plot with one variable
`geom_density2d(...)` | Density plot with two variables
`geom_histogram(...)` | Histogram

#### A bar chart

```{r}
qplot(x = race, data = birthwt, geom = "bar")
```

#### Histograms and density plots

```{r, fig.width=12, fig.height=6, dpi=70, cache=TRUE, eval=FALSE}
base.plot <- ggplot(birthwt, aes(x = mother.age)) +
  xlab("Mother's age") 
base.plot + geom_histogram()
base.plot + geom_histogram(aes(fill = race), alpha = 0.5)

base.plot <- ggplot(birthwt, aes(x = mother.age)) +
  xlab("Mother's age") 
base.plot + geom_density()
base.plot + geom_density(aes(fill = race), alpha = 0.5)

```

#### Box plots and violin plots

```{r}
base.plot <- ggplot(birthwt, aes(x = as.factor(physician.visits), y = birthwt.grams)) +
  xlab("Number of first trimester physician visits") +
  ylab("Baby's birthweight (grams)")
# Violin plot
base.plot + geom_violin()
```

#### Visualizing means

Previously we calculated the following table:
```{r}
bwt.summary <- aggregate(birthwt.grams ~ race + mother.smokes, data = birthwt, FUN = mean) # aggregate
bwt.summary
```

We can plot this table in a nice bar chart as follows:

```{r, fig.height=5, fig.width=5, fig.align='center', eval=TRUE}
# Define basic aesthetic parameters
p.bwt <- ggplot(data = bwt.summary, aes(y = birthwt.grams, x = race, fill = mother.smokes))

# Pick colors for the bars
bwt.colors <- c("#009E73", "#999999")

# Display barchart
p.bwt + geom_bar(stat = "identity", position = "dodge") +
  ylab("Average birthweight") + 
  xlab("Mother's race") +
  guides(fill = guide_legend(title = "Mother's smoking status")) + 
  scale_fill_manual(values=bwt.colors)
```

#### Does the association between birthweight and mother's age depend on smoking status?

We previously ran the following command to calculate the correlation between mother's ages and baby birthweights.

```{r}
by(data = birthwt[c("birthwt.grams", "mother.age")], 
   INDICES = birthwt["mother.smokes"], 
   FUN = function(x) {cor(x[,1], x[,2])})
```

Here's a visualization of our data that allows us to see what's going on.

```{r, fig.height=5, fig.width=6, fig.align='center', eval=TRUE}
ggplot(birthwt, aes(x=mother.age, y=birthwt.grams, shape=mother.smokes, color=mother.smokes)) + 
  geom_point() + # Adds points (scatterplot)
  geom_smooth(method = "lm") + # Adds regression lines
  ylab("Birth Weight (grams)") + # Changes y-axis label
  xlab("Mother's Age (years)") + # Changes x-axis label
  ggtitle("Birth Weight by Mother's Age") # Changes plot title
```