
###Agenda

- Writing functions in R
- If-else statements

###Basics of lists

> A list is a **data structure** that can be used to store **different kinds** of data

- Recall: a vector is a data structure for storing *similar kinds of data*

- To better understand the difference, consider the following example.

```{r}
my.vector.1 <- c("Ela", 22, FALSE) # (name, weight, is.male)
my.vector.1 
typeof(my.vector.1)  # All the elements are now character strings!
```

###Lists vs. vectors

```{r}
#Info about Lara 
my.vector.2 <- c(FALSE, TRUE, 1) # (is.male, is.baby, age)
typeof(my.vector.2)
```

- Vectors expect elements to be all of the same type (e.g., `Boolean`, `numeric`, `character`)

- When data of different types are put into a vector, the R converts everything to a common type

###Lists

- To store data of different types in the same object, we use lists

- Simple way to build lists: use `list()` function

```{r}
my.list <- list("Ela", 22, TRUE)
my.list
sapply(my.list, typeof)
```

###Named elements
```{r}
kid.1 <- list(name="Ela", weight=22, is.male=FALSE)
kid.1
```

###Referencing elements of a list (similar to data frames)
```{r}
kid.1$name # Get "name" element (returns a string)
kid.1[["name"]] # Get "name" element (returns a string)
kid.1["name"] # Get "name" slice (returns a sub-list)
c(typeof(kid.1$name), typeof(kid.1["name"]))
```

###Functions
- We have used a lot of built-in functions: `mean()`, `subset()`, `plot()`, `read.table()`...

- An important part of programming and data analysis is to write custom functions

- Functions help make code **modular**

- Functions make debugging easier

- Remember: this entire class is about applying *functions* to *data*

###What is a function?

> A function is a machine that turns **input objects** (arguments) into an **output object** (return value) according to a definite rule.

- Let's look at a really simple function

```{r}
addOne <- function(x) {
  x + 1
}
```

- `x` is the **argument** or **input**

- The function **output** is the input `x` incremented by 1

```{r}
addOne(12)
```

###More interesting example

- Here's a function that returns a % given a numerator, denominator, and desired number of decimal values

```{r}
calculatePercentage <- function(x, y, d) {
  decimal <- x / y  # Calculate decimal value
  round(100 * decimal, d)  # Convert to % and round to d digits
}

calculatePercentage(27, 80, 1)
```

- If you're calculating several %'s for your report, you should use this kind of function instead of repeatedly copying and pasting code

###Function returning a list
- Here's a function that takes a person's full name (FirstName LastName), weight in lb and height in inches and converts it into a list with the person's first name, person's last name, weight in kg, height in m, and BMI.

```{r}
createPatientRecord <- function(full.name, weight, height) {
  name.list <- strsplit(full.name, split=" ")[[1]]
  first.name <- name.list[1]
  last.name <- name.list[2]
  weight.in.kg <- weight / 2.2
  height.in.m <- height * 0.0254
  bmi <- weight.in.kg / (height.in.m ^ 2)
  list(first.name=first.name, last.name=last.name, weight=weight.in.kg, height=height.in.m,
       bmi=bmi)
}
```
###Trying out the function
```{r}
createPatientRecord("Ahmet CAN", 185, 12 * 6 + 1)
```

###Another example: 5 number summary
- Calculate mean, 10% trimmed mean, median, geometric and harmonic means of a random variable 

```{r}
fiveaverages <- function(x) {
  c(average=mean(x), trimmed=mean(x, trim=0.10), median=median(x), geometricmean= prod(x)^(1/length(x)), harmonicmean= 1/mean(1/x))
}
x <- rnorm(200, mean=20, sd=5) # Vector of 200 kids with weight mean 5 and weight sd 2
par(mfrow=c(1,3))
hist(x)
qqnorm(x)
qqline(x)
boxplot(x)
fiveaverages(x)
```
###Add outliers in the dataset
- Calculate mean, 10% trimmed mean, median, geometric and harmonic means of a random variable 

```{r}
x <- c(rnorm(180, mean=20, sd=5),rnorm(20, mean=50, sd=5)) # Vector of 200 kids with outliers
par(mfrow=c(1,3))
hist(x)
qqnorm(x)
qqline(x)
boxplot(x)
fiveaverages(x)
```
###If-else statements

- Oftentimes we want our code to have different effects depending on the features of the input

- Example: Calculating a student's letter grade
  - If grade >= 90, assign A
  - Otherwise, if grade >= 80, assign B
  - Otherwise, if grade >= 70, assign C
  - In all other cases, assign F

- To code this up, we use if-else statements

###If-else Example: Letter grades
```{r}
calculateLetterGrade <- function(x) {
  if(x >= 90) {
    grade <- "A"
  } else if(x >= 80) {
    grade <- "B"
  } else if(x >= 70) {
    grade <- "C"
  } else {
    grade <- "F"
  }
  grade
}

course.grades <- c(92, 78, 87, 91, 62)
sapply(course.grades, FUN=calculateLetterGrade)
```

###`return()`

- In the previous examples we specified the output simply by writing the output variable as the last line of the function

- More explicitly, we can use the `return()` function

```{r}
addOne <- function(x) {
  return(x + 1)
}

addOne(12)
```

- We will generally avoid the `return()` function, but you can use it if necessary or if it makes writing a particular function easier.

###Write your own function for Median Absolute Deviation
```{r,eval=FALSE}
MADcalculator <- function(x) {

  ...
  ...
  ...
  
  
}
```
###Write your own function for correlation analysis

- Example: 
  - Draw a scatter plot for x and y
  - Calculate Pearson corelation coefficient
  - Automatically evaluate if the relationship is strong, weak, etc...
  
```{r,eval=FALSE}
CorrelationAnalyzer <- function(x,y) {

  ...
  ...
  ...
  
  
}
```

###Write your own function for replacing missing values with the 10% trimmed mean 

```{r,eval=FALSE}
MissingValueReplacer <- function(x,y) {

  ...
  ...
  ...
  
  
}
```